---
title: "Final Code"
author: "Thomas Stranick"
date: "2025-12-12"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggplot2)
library(stringr)
```

## Data Pre-Processing

Read in Excel, convert to tibble for ease of use
Select only SAs with no dependents
Getting rows with premium data, do not want to include NA, or other non-values

```{r}
  survey_data <- read.csv("Data/adult24.csv")
  survey_tibble <- tibble(survey_data)

  total_surveyed = nrow(survey_tibble)
  
  survey_singles <- survey_tibble %>% filter(PRPLCOV1_A == 2)
  
  invalid_premium <- c(99999, 99998, 99997)

  premium_data <- survey_singles %>% 
  filter(!is.na( HICOSTR1_A ) & !HICOSTR1_A %in% invalid_premium) 
```

Select covariates to explore

```{r}
premium_data <- premium_data %>% 
  select(HICOSTR1_A, #response y
        AGEP_A, SEX_A, RACEALLP_A, HISPALLP_A, MARITAL_A, NATUSBORN_A, ORIENT_A, #demographic
        POVRATTC_A, RATCAT_A, EDUCP_A, ACCSSINT_A, #socio-economic
        DIBEV_A, PREDIB_A, HYPEV_A, ASEV_A, EVERCOVD_A, ANXFREQ_A,DEPFREQ_A, #health conditions
        SMOKELSEV1_A, SMKCIGST_A, DRKLIFE_A, DRKSTAT_A, MODFREQW_A, #health related behaviors
        REGION, URBRRL23, #geography
        PPSU, PSTRAT, WTFA_A)#survey design
```

making invalid values NA (refused, not ascertained, don't know)

```{r}
premium_data$AGEP_A[premium_data$AGEP_A %in% c(97, 98, 99)] <- NA
premium_data$SEX_A[premium_data$SEX_A %in% c(7,8,9)] <- NA
premium_data$RACEALLP_A[premium_data$RACEALLP_A %in% c(7,8,9)] <- NA
premium_data$HISPALLP_A[premium_data$HISPALLP_A %in% c(97,98,99)] <- NA
premium_data$MARITAL_A[premium_data$MARITAL_A %in% c(7,8,9)] <- NA
premium_data$NATUSBORN_A[premium_data$NATUSBORN_A %in% c(7,8,9)] <- NA
premium_data$ORIENT_A[premium_data$ORIENT_A %in% c(7,8)] <- NA
premium_data$RATCAT_A[premium_data$RATCAT_A %in% c(98)] <- NA
premium_data$EDUCP_A[premium_data$EDUCP_A %in% c(97,98,99)] <- NA
premium_data$ACCSSINT_A[premium_data$ACCSSINT_A %in% c(7,8,9)] <- NA
premium_data$NATUSBORN_A[premium_data$NATUSBORN_A %in% c(7,8,9)] <- NA
premium_data$DIBEV_A[premium_data$DIBEV_A %in% c(7,8,9)] <- NA
premium_data$PREDIB_A [premium_data$PREDIB_A %in% c(7,8,9)] <- NA
premium_data$ASEV_A[premium_data$ASEV_A %in% c(7,8,9)] <- NA
premium_data$EVERCOVD_A[premium_data$EVERCOVD_A %in% c(7,8,9)] <- NA
premium_data$ANXFREQ_A[premium_data$ANXFREQ_A %in% c(7,8,9)] <- NA
premium_data$DEPFREQ_A[premium_data$DEPFREQ_A %in% c(7,8,9)] <- NA
premium_data$SMOKELSEV1_A[premium_data$SMOKELSEV1_A %in% c(7,8,9)] <- NA
premium_data$SMKCIGST_A[premium_data$SMKCIGST_A %in% c(9)] <- NA
premium_data$DRKLIFE_A[premium_data$DRKLIFE_A %in% c(7,8,9)] <- NA
premium_data$DRKSTAT_A[premium_data$DRKSTAT_A %in% c(10)] <- NA
premium_data$MODFREQW_A[premium_data$MODFREQW_A %in% c(97,98,99)] <- NA
```

Convert categorical variables to factors

```{r}
premium_data <- premium_data %>% 
  mutate(across(c(SEX_A, RACEALLP_A, HISPALLP_A, MARITAL_A, NATUSBORN_A, ORIENT_A,
                  RATCAT_A, EDUCP_A, ACCSSINT_A,
                  DIBEV_A, PREDIB_A, HYPEV_A, ASEV_A, EVERCOVD_A, ANXFREQ_A,DEPFREQ_A,
                  SMOKELSEV1_A, SMKCIGST_A, DRKLIFE_A, DRKSTAT_A, MODFREQW_A,
                  REGION, URBRRL23),as.factor))
```


## Numerical and Graphical Summary of Data

```{r}
str(premium_data)
summary(premium_data)

num_total <- nrow(survey_data)

num_sample_adults <- nrow(premium_data) #number of SA with premium value

num_obs <- nobs(premium_model) #number of observations in model

num_NAs <- num_sample_adults - num_obs #number of people who did not answer some of the questions

sample_table <- data.frame(
  total_adults = num_total,
  premium_adults = num_sample_adults,
  invalid_responses = num_NAs
)

sample_table
```

##Premiums

```{r}
p <- ggplot(premium_data, aes(x = HICOSTR1_A)) +
  geom_histogram(bins = 50, fill = "skyblue", color = "black")

# Extract histogram data
hist_data <- ggplot_build(p)$data[[1]]

# Find tallest bin
top_bin <- hist_data %>%
  slice_max(ncount, n = 1)   # or 'count' if you prefer raw counts

# Add label to the tallest bin
p + 
  annotate("text",
           x = top_bin$x,          # bin center
           y = top_bin$count,      # height of bar
           label = paste0("Peak: ", round(top_bin$x, 2)),
           vjust = -0.5,
           color = "black",
           fontface = "bold")
```

##Demographics 
```{r}
#Age Histogram
ggplot(premium_data, aes(x = AGEP_A)) +
  geom_histogram(
    binwidth = 5,
    fill = "lightblue",
    color = "black",
    size = 1,          # thicker bar borders
    alpha = 0.7        # more solid bars
  ) +
  theme_minimal() +
  labs(
    title = "Age Distribution",
    x = "Age",
    y = "Count"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"), # centered title
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 12)
  )

#Gender Table
gender_table <- table(premium_data$SEX_A)
names(gender_table) <- c("Male", "Female")
gender_table

#Race Histogram

freq_table <- table(premium_data$HISPALLP_A)

df <- as.data.frame(freq_table)
names(df) <- c("Code", "Frequency")

labels <- c(
  "1" = "Hispanic",
  "2" = "Non-Hispanic White only",
  "3" = "Non-Hispanic Black/African American only",
  "4" = "Non-Hispanic Asian only",
  "5" = "Non-Hispanic AIAN only",
  "6" = "Non-Hispanic AIAN and any other group",
  "7" = "Other single and multiple races")

df$Description <- str_wrap(labels[as.character(df$Code)], width = 20)

ggplot(df, aes(x = reorder(Description, Frequency), y = Frequency)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +
  geom_text(aes(label=Frequency), hjust=-0.1) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.2))) +
  theme_minimal() +
  labs(title = "Frequency of Race/Ethnicity Groups",
       x = "Race/Ethnicity Group",
       y = "Frequency") +
  theme(plot.title = element_text(hjust = 0.5))

```

## Socio-Economic
```{r}
#Poverty Ratio Density Plot

ggplot(premium_data, aes(x = POVRATTC_A)) +
  geom_density(
    fill = "lightgreen",
    alpha = 0.6,
    color = "darkgreen",
    size = 1.1          # thicker, more visible curve
  ) +
  theme_minimal() +
  labs(
    title = "Family Poverty Ratio",
    x = "Poverty Ratio",
    y = "Density"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 12)
  )
```

## Health Totals
```{r}
#DIBEV_A + PREDIB_A + HYPEV_A + ASEV_A + ANXFREQ_A + DEPFREQ_A + #health conditions

dibev_yes <- sum(premium_data$DIBEV_A == 1, na.rm = TRUE)
dibev_no  <- sum(premium_data$DIBEV_A == 2, na.rm = TRUE)

predib_yes <- sum(premium_data$PREDIB_A == 1, na.rm = TRUE)
predib_no  <- sum(premium_data$PREDIB_A == 2, na.rm = TRUE)

hypev_yes <- sum(premium_data$HYPEV_A == 1, na.rm = TRUE)
hypev_no  <- sum(premium_data$HYPEV_A == 2, na.rm = TRUE)

asev_yes <- sum(premium_data$ASEV_A == 1, na.rm = TRUE)
asev_no  <- sum(premium_data$ASEV_A == 2, na.rm = TRUE)

health_conditions <- data.frame(
  Condition = c("Diabetes", "Prediabetes", "Hypertension", "Asthma"),
  Has = c(dibev_yes, predib_yes, hypev_yes, asev_yes),
  Does_Not  = c(dibev_no,  predib_no,  hypev_no,  asev_no)
)

health_conditions
```




